Syntax and Semantics of Cedille

### 2021-01-02 docs/semantics/paper.tex
docs: semantics:
- make rho checking
- give full listing of conversion rules
- minor revisions to logical consistency, normalization theorem

\usepackage{amsmath,amssymb,amsthm}
\usepackage{url}
\usepackage{comment}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{parskip}
\usepackage{fullpage}

@@

Call-by-name reduction is \emph{not} what underpins the
congruence relation \(\cong\) for types or terms, which is full
\(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms)
modulo erasure of annotations in terms.
The erasure operation is defined in Figure~\ref{fig:eraser}, and is essentially
the \emph{extraction} function for the Implicit Calculus of Constructions given
by~\cite{BB08_ICC-as-a-Programming-Language} adapted to CDLE.
To understand the role of erasure, recall that the type theory is \emph{extrinsic} (a.k.a.\
Curry-style), and hence we only consider erasures \(|t|\) of terms when testing
convertibility.
This is lifted to the conversion relation on types \(T \cong T'\) and kinds
\(\kappa \cong \kappa'\), whose central rules are given in
Figure~\ref{fig:conv}.
That figure omits the various congruence rules needed to equate bigger
expressions by equating subexpressions (the convertibility rules for kinds
consists entirely of congruence rules).

@@

    \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
          {
           \begin{array}{cc}
             T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
             & \Gamma \vdash t_1 \tpcheck T_1
             \\ \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
             & |t_1| =_{\beta\eta} |t_2|
           \end{array}
          } &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x] T'}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
          {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}
          {\Gamma\vdash t\tpcheck \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}} &
    \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
          {
          \begin{array}{ccc}
            \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
            & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
            & |t_2'| =_{\beta\eta} |t_2|
            \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
            & \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {
              \Gamma\vdash t\tpcheck \{t' \simeq t''\}
              & \Gamma\vdash t' \Leftrightarrow T
              & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
          }
  \end{array}
    \end{array}
  \]
\caption{Rules for checking a term against a type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{lllllll}
    |x| & = & x &\ &
    |\absu{\lambda}{x}{t}| & = & \absu{\lambda}{x}{|t|} \\
    |t\ t'| & = & |t|\ |t'| &\ &
    |t\cdot T| & = & |t| \\
    |\absu{\Lambda}{x}{t}| & = & |t| &\ &
    |t\ \mhyph t'| & = & |t| \\
    |[t , t']| & = & |t| &\ &
    |t.1| & = & |t| \\
    |t.2| & = & |t| &\ &
    |\beta\{t\}| & = & |t|\\
    |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
    |\rho\ t\ @x\langle t_2 \rangle.T\ \mhyph\ t'| & = & |t'| \\
    |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
    |\chi\ T\ \mhyph\ t'| & = & |t'|
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{T \cong T'}{T \leadsto^*_\beta T_1 & T' \leadsto^*_\beta T_2 & T_1\cong^t T_2}  & 
    \infer{T \cong T'}{T \cong^t T'} \\ \\
    \infer{T\ t \cong^t T\ t'}{T \cong^t T' & |t| =_{\beta\eta} |t'|} &
    \infer{\{ t_1 \simeq t_2 \} \cong^t \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
  \end{array}
  \]
  \caption{Non-congruence rules for conversion}
  \label{fig:conv}
\end{figure}

@@

  The Kleene trick means that in Cedille, any such term --- even otherwise untypable
  terms, non-normalizing terms, etc. --- prove trivially true equations.
  Put another way, any trivially true equation type in CDLE is a suitable type
  to classify all untyped lambda calculus terms.
\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
  \(t_2'\) in the synthesized type of its second subexpression.
  The version presented here requires a type annotations (``guide'') \(@x\langle
  t_2 \rangle.T\), where \(x\) indicates the occurrences of \(t_1\) to
  substitute and \(t_2\) is a term, \(\beta\eta\)-equivalent to \(t_2'\), that
  is used to replace \(x\).
  There is no requirement that \(t_1\) and \(t_2\) will have the same type, or
  even be typable.
  However, \([t_2/x]\ T\) must be a well-kinded type and \([t_1/x]T\) (which
  might be ill-kinded) must be convertible with \(T'\) modulo erasure.
  In Cedille, the guide is optional and the construct may be used to rewrite a
  contextually given type; a heuristic, whose details are beyond the scope of
  this document, is used to produce a resulting type that is well-kinded.

@@

Figure~\ref{fig:semtp} gives a realizability semantics for types and
kinds, following the semantics given in the previous papers on
CDLE~\cite{stump18,stump17}.  Details of this semantics are presented
further in Section~\ref{sec:snd} below.  Using the semantics and the
definition in Figure~\ref{fig:semctxt} of $\interp{\Gamma}$, we can
prove the following theorem:
\begin{theorem}[Soundness]
\label{thm:snd}
Suppose $(\sigma,\rho)\in\interp{\Gamma}$.  Then we have:
\begin{enumerate}
\item If $\Gamma\vdash \kappa$, then $\interp{\kappa}_{\sigma,\rho}$ is defined.
\item If $\Gamma\vdash T \tpsynth \kappa$, then $\interp{T}_{\sigma,\rho}\in\interp{\kappa}_{\sigma,\rho}$.
\item If $\Gamma\vdash t \tpsynth T$ then $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $\Gamma\vdash t \tpcheck T$ and $\interp{T}_{\sigma,\rho}\in \mathcal{R}$, then
    $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $T \cong T'$ or $T \cong^t T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
\item If \(\kappa \cong \kappa'\) and \(\interp{\kappa}_{\sigma,\rho}\) and
  \(\interp{\kappa'}_{\sigma,\rho}\) are both defined, then they are equal.
\end{enumerate}
\end{theorem}

@@

\begin{theorem}[Logical consistency]
\label{thm:consis}
  There is no term $t$ such that $\cdot \vdash t : \abs{\forall}{X}{\star}{X}$.
\end{theorem}

@@

\begin{theorem}[Call-by-name normalization of functions]
  \label{thm:cedille-termination}
  Suppose \(\cdot \vdash t \tpsynth T\) and there exists a closed term \(t'\)
  such that \(\cdot \vdash t' \tpsynth T \to \abs{\Pi}{x}{T_1}{T_2}\) for some
  \(T_1,T_2\) and \(|t'| = \absu{\lambda}{x}{x}\).
  Then \(|t|\) is call-by-name normalizing.
\end{theorem}

@@

\startcase{.2cm}
\[
  \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
  {
    \begin{array}{ccc}
      \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
      & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
      & |t_2'| =_{\beta\eta} |t_2|
      \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
      & \Gamma \vdash t \tpsynth T'
      & T' \cong [t_1/x]T
    \end{array}
  }
\]
By the IH applied to the first premise in the first row, $\sigma|t_1| =_{\cbe}
\sigma|t_2'|$.
With the second and third premise, and from the assumption \(\sigma \in
\interp{\Gamma}\), we have \(\sigma|t_2'| =_{\cbe} \sigma|t_2|\) (and
\(\sigma|t_2|\) is closed), so we obtain
\([\sigma|t_1|]_{\cbe} = [\sigma|t_2|]_{\cbe}\).
By the IH applied to the first premise of the second row,
\(\interp{[t_2/x]T}_{\sigma,\rho} \in \mathcal{R}\) and so is defined.
By the IH applied to the first premise in the second row,
\([\sigma|t|]_{c\beta\eta} \in \interp{T'}_{\sigma,\rho}\).
By the IH applied to the second premise in the second row,
\(\interp{T'}_{\sigma,\rho} = \interp{[t_1/x]T}_{\sigma,\rho}\).
The result then follows by applying Lemma~\ref{lem:termsubstinterp}.

@@

\subsection*{Proof of part (5) }

\startcase{.2cm}
\[
    \infer{T \cong T'}{T \leadsto^*_\beta T_1 & T' \leadsto^*_\beta T_2 & T_1\cong^t T_2}  
\]
By Lemma~\ref{lem:interppres}, we have
\[
\begin{array}{lll}
  \interp{T}_{\sigma,\rho} & = & \interp{T_1}_{\sigma,\rho}\\
  \interp{T'}_{\sigma,\rho} & = & \interp{T_2}_{\sigma,\rho}
\end{array}
\]
By the IH for the third premise, we have $\interp{T_1}_{\sigma,\rho} = \interp{T_2}_{\sigma,\rho}$,
which suffices.

\startcase{.2cm}
\[
    \infer{T \cong T'}{T \cong^t T'} 
\]
By the IH.

\startcase{.2cm}
\[
    \infer{T\ t \cong^t T'\ t'}{T \cong^t T' & |t| =_{\beta\eta} |t'|}
\]
By the semantics, $\interp{T\ t}_{\sigma,\rho} =
\interp{T}_{\sigma,\rho}([\sigma|t|]_{\cbe})$.  By the second premise
and the IH for the first premise, this equals
$\interp{T'}_{\sigma,\rho}([\sigma|t'|]_{\cbe})$, as required.

\startcase{.2cm}
\[
    \infer{\{ t_1 \simeq t_2 \} \cong^t \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
\]
This follows easily from the premises and the semantics of equality types.

@@

@@

  \startcase{.2cm}
  \[
    \infer{
      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \cdot T' \tpsynth [T'/X]T
    }{
      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads
      \abs{\forall}{X}{\kappa}{T}
      & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T' \tpsynth \kappa'
      & \kappa'\cong\kappa
    }
  \]
  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t
  \tpsynthleads \abs{\forall}{X}{\kappa}{T}\).
  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T'
  \tpsynth \kappa''\) for some \(\kappa'' \cong \kappa'\), from which we obtain
  \(\kappa'' \cong \kappa\).
  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \cdot T' \tpsynth
  [T'/X]T\), and the measure is preserved.

  \startcase{.2cm}
  \[
    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
    {
      \begin{array}{ccc}
        \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
        & |t_2'| =_{\beta\eta} |t_2|
        \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star
        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynth T'
        & T' \cong [t_1/x]T
      \end{array}
    }
  \]
  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t'
  \tpsynthleads \{t_1 \simeq t_2'\}\).
  By the IH on the second premise of the second row,
  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star\)
  (\(\star\) is only convertible with itself).
  We apply the IH to the second premise of the second row and use the rule to conclude
  (\(\textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2) =
  \textit{dom}(\Gamma_1,X_1:\kappa_1',\Gamma_2)\)).
\end{proof}

@@

>>>>

\usepackage{amsmath,amssymb,amsthm,MnSymbol}
\usepackage{url}
\usepackage{comment}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{parskip}
\usepackage{fullpage}

@@

Call-by-name reduction is \emph{not} what underpins the
equivalence relation \(\cong\) for types and kinds, which is full
\(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms)
modulo erasure of annotations in terms.
The erasure operation is defined in Figure~\ref{fig:eraser}, and is essentially
the \emph{extraction} function for the Implicit Calculus of Constructions given
by~\cite{BB08_ICC-as-a-Programming-Language} adapted to CDLE.
To understand the role of erasure, recall that the type theory is \emph{extrinsic} (a.k.a.\
Curry-style), and hence we only consider erasures \(|t|\) of terms when testing
convertibility.
This is lifted to the conversion relation on types \(T \cong T'\) and kinds
\(\kappa \cong \kappa'\), whose rules are given in
Figure~\ref{fig:conv}.

@@

    \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
          {
           \begin{array}{cc}
             T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
             & \Gamma \vdash t_1 \tpcheck T_1
             \\ \Gamma \vdash t_2 \tpcheck [t_1/x]^{T_1}\ T_2
             & |t_1| =_{\beta\eta} |t_2|
           \end{array}
          } &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x] T'}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
          {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}
          {\Gamma\vdash t\tpcheck \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}} &
    \infer{\Gamma \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T}
          {
          \begin{array}{ccc}
            \Gamma \vdash t \tpsynthleads \{t_1' \simeq t_2\}
            & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma)
            & |t_1'| =_{\beta\eta} |t_1|
            \\ \Gamma \vdash [t_1/x] T' \tpsynth \star
            & \Gamma \vdash t' \tpcheck [t_1/x]T'
            & [t_2/x]T' \cong T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {
              \Gamma\vdash t\tpcheck \{t' \simeq t''\}
              & \Gamma\vdash t' \Leftrightarrow T
              & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
          }
  \end{array}
    \end{array}
  \]
\caption{Rules for checking a term against a type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{lllllll}
    |x| & = & x &\ &
    |\absu{\lambda}{x}{t}| & = & \absu{\lambda}{x}{|t|} \\
    |t\ t'| & = & |t|\ |t'| &\ &
    |t\cdot T| & = & |t| \\
    |\absu{\Lambda}{x}{t}| & = & |t| &\ &
    |t\ \mhyph t'| & = & |t| \\
    |[t , t']| & = & |t| &\ &
    |t.1| & = & |t| \\
    |t.2| & = & |t| &\ &
    |\beta\{t\}| & = & |t|\\
    |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
    |\rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t'| & = & |t'| \\
    |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
    |\chi\ T\ \mhyph\ t'| & = & |t'|
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

\begin{figure}
  \[
    \begin{array}{c}
      \fbox{\(T_1 \cong T_2\)}
      \quad \fbox{\(T_1 \cong^{\text{t}} T_2\)}
      \\ \\
      \infer{
       T_1 \cong T_2
      }{
       T_1 \leadsto^*_\beta T_1' \nleadsto_\beta
       \quad T_2 \leadsto^*_\beta T_2' \nleadsto_\beta
       \quad T_1' \cong^{\text{t}} T_2'
      }
      \\ \\
      \begin{array}{cc}
        \infer{X \cong^{\text{t}} X}{}
        &
        \infer{
         \abs{\forall}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\forall}{X}{\kappa_2}{T_2}
        }{
         \kappa_1 \cong \kappa_2
         \quad  T_1 \cong T_2
        }
        \\ \\
          \infer{
           \abs{\forall}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\forall}{x}{T_2}{T_2'}
          }{
           T_1 \cong T_2
           \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         \abs{\Pi}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\Pi}{x}{T_2}{T_2'}
        }{
         T_1 \cong T_2
         \quad  T_1' \cong T_2'
        }
        \\ \\ 
          \infer{
           \abs{\lambda}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\lambda}{x}{T_2}{T_2'}
          }{
           T_1 \cong T_2
           \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         \abs{\lambda}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\lambda}{X}{\kappa_2}{T_2}
        }{
         \kappa_1 \cong \kappa_2
         \quad  T_1 \cong T_2
        }
        \\ \\
          \infer{
          \abs{\iota}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\iota}{x}{T_2}{T_2'}
          }{
          T_1 \cong T_2
          \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         T_1\ t_1 \cong^{\text{t}} T_2\ t_2
        }{
         T_1 \cong^{\text{t}} T_2 \quad |t_1| =_{\beta\eta} |t_2|
        }
        \\ \\
          \infer{
           T_1 \cdot T_1' \cong^{\text{t}} T_2 \cdot T_2'
          }{
           T_1 \cong^{\text{t}} T_2
           \quad T_1' \cong T_2'
          }
        &
        \infer{
         \{ t_1 \simeq t_2 \} \cong^{\text{t}} \{ t_1'\ \simeq t_2' \}
        }{
         |t_1| =_{\beta\eta} |t_1'| \quad |t_2| =_{\beta\eta} |t_2'|
        }
      \end{array}
      \\ \\ \fbox{\(\kappa_1 \cong \kappa_2\)}
      \\ \\
      \begin{array}{cc}
        \infer{
         \star \cong \star
        }{}
        & \infer{
           \abs{\Pi}{x}{T_1}{\kappa_1} \cong \abs{\Pi}{x}{T_2}{\kappa_2}
          }{
           T_1 \cong T_2
           \quad \kappa_1 \cong \kappa_2
          }
        \\ \\
        \infer{
         \abs{\Pi}{X}{\kappa_1}{\kappa_1'} \cong \abs{\Pi}{X}{\kappa_2}{\kappa_2'}
        }{
         \kappa_1 \cong \kappa_2
         \quad \kappa_1' \cong \kappa_2'
        }
      \end{array}
    \end{array}
  \]
  \caption{Conversion rules for classifiers}
  \label{fig:conv}
\end{figure}

@@

  The Kleene trick means that in Cedille, any such term --- even otherwise untypable
  terms, non-normalizing terms, etc. --- prove trivially true equations.
  Put another way, any trivially true equation type in CDLE is a suitable type
  to classify all untyped lambda calculus terms.
\item The \(\rho\) construct allows one to rewrite occurrences of \(t_2\) in the
  checked type of the \(T\) of the whole expression to \(t_1\) before checking
  the type of the subexpression \(t'\).
  The version presented here requires a type annotations (``guide'') \(@x\langle
  t_1 \rangle.T\).
  The first subexpression \(t\) must synthesize (possibly after some
  normalization) an equation type \(\{t_1' \simeq t_2\}\), and the user provided
  term \(t_1\) must be \(\beta\eta\)-convertible (modulo erasure) to \(t_1'\).
  Then, the user provided type \(T'\) is checked to have kind \(\star\) after
  replacing occurrences of \(x\) with \(t_1\), and the subexpression \(t'\) is
  checked against this type.
  Finally, \([t_2/x]T'\) (which need not be well-kinded) must be convertible
  with the expected type \(T\).
  In Cedille, the guide is optional and the construct may be used to rewrite a
  contextually given type; a heuristic, whose details are beyond the scope of
  this document, is used to produce a resulting type that is well-kinded.

@@

Figure~\ref{fig:semtp} gives a realizability semantics for types and
kinds, following the semantics given in the previous papers on
CDLE~\cite{stump18,stump17}.  Details of this semantics are presented
further in Section~\ref{sec:snd} below.  Using the semantics and the
definition in Figure~\ref{fig:semctxt} of $\interp{\Gamma}$, we can
prove the following theorem:
\begin{theorem}[Soundness]
\label{thm:snd}
Suppose $(\sigma,\rho)\in\interp{\Gamma}$.  Then we have:
\begin{enumerate}
\item If $\Gamma\vdash \kappa$, then $\interp{\kappa}_{\sigma,\rho}$ is defined.
\item If $\Gamma\vdash T \tpsynth \kappa$, then $\interp{T}_{\sigma,\rho}\in\interp{\kappa}_{\sigma,\rho}$.
\item If $\Gamma\vdash t \tpsynth T$ then $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $\Gamma\vdash t \tpcheck T$ and $\interp{T}_{\sigma,\rho}\in \mathcal{R}$, then
    $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $T \cong T'$ or $T \cong^{\text{t}} T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
\item If \(\kappa \cong \kappa'\) and \(\interp{\kappa}_{\sigma,\rho}\) and
  \(\interp{\kappa'}_{\sigma,\rho}\) are both defined, then they are equal.
\end{enumerate}
\end{theorem}

@@

\begin{theorem}[Logical consistency]
\label{thm:consis}
  There is no term $t$ such that $\cdot \vdash t \tpsynth \abs{\forall}{X}{\star}{X}$.
\end{theorem}

@@

\begin{theorem}[Call-by-name normalization of functions]
  \label{thm:cedille-termination}
  Suppose \(\cdot \vdash t \tpsynth T\) and \(\cdot \vdash t' \tpsynth T \to
  \abs{\Pi}{x}{T_1}{T_2}\), and furthermore that \(|t'| =
  \absu{\lambda}{x}{x}\).
  Then \(|t|\) is call-by-name normalizing.
\end{theorem}

@@

\startcase{.2cm}
\[
  \infer{\Gamma \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T}
  {
    \begin{array}{ccc}
      \Gamma \vdash t \tpsynthleads \{t_1' \simeq t_2\}
      & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma)
      & |t_1'| =_{\beta\eta} |t_1|
      \\ \Gamma \vdash [t_1/x] T' \tpsynth \star
      & \Gamma \vdash t' \tpcheck [t_1/x]T'
      & [t_2/x]T' \cong T
    \end{array}
  }
\]
By the IH applied to the first premise in the first row, $\sigma|t_1'| =_{\cbe}
\sigma|t_2|$.
With the second and third premise, and from the assumption \(\sigma \in
\interp{\Gamma}\), we have \(\sigma|t_1'| =_{\cbe} \sigma|t_1|\) (and
\(\sigma|t_1|\) is closed), so we obtain
\([\sigma|t_1|]_{\cbe} = [\sigma|t_2|]_{\cbe}\).
By the IH applied to the first premise of the second row,
\(\interp{[t_1/x]T}_{\sigma,\rho} \in \mathcal{R}\) and so is defined.
By the IH applied to the second premise in the second row,
\([\sigma|t'|]_{c\beta\eta} \in \interp{[t_1/x]T'}_{\sigma,\rho}\).
By the IH applied to the second premise in the second row,
\(\interp{T}_{\sigma,\rho} = \interp{[t_2/x]T'}_{\sigma,\rho}\).
The result then follows by applying Lemma~\ref{lem:termsubstinterp}.

@@

\subsection*{Proof of part (5) }
We show the cases for the non-congruential rules of Figure~\ref{fig:conv}.

\startcase{.2cm}
\[
  \infer{
    T_1 \cong T_2
  }{
    T_1 \leadsto^*_\beta T_1' \nleadsto_\beta
    \quad T_2 \leadsto^*_\beta T_2' \nleadsto_\beta
    \quad T_1' \cong^{\text{t}} T_2'
  }
\]
By Lemma~\ref{lem:interppres}, we have
\[
\begin{array}{lll}
  \interp{T_1}_{\sigma,\rho} & = & \interp{T_1'}_{\sigma,\rho}\\
  \interp{T_2}_{\sigma,\rho} & = & \interp{T_2'}_{\sigma,\rho}
\end{array}
\]
By the IH for the third premise, we have $\interp{T_1'}_{\sigma,\rho} = \interp{T_2'}_{\sigma,\rho}$,
which suffices.

\startcase{.2cm}
\[
    \infer{T_1\ t_1 \cong^{\text{t}} T_2\ t_2}{T_1 \cong^{\text{t}} T_1 & |t_1| =_{\beta\eta} |t_2|}
\]
By the semantics, $\interp{T_1\ t_1}_{\sigma,\rho} =
\interp{T_1}_{\sigma,\rho}([\sigma|t_1|]_{\cbe})$.  By the second premise
and the IH for the first premise, this equals
$\interp{T_2}_{\sigma,\rho}([\sigma|t_2|]_{\cbe})$, as required.

\startcase{.2cm}
\[
    \infer{\{ t_1 \simeq t_2 \} \cong^{\text{t}} \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
\]
This follows easily from the premises and the semantics of equality types.

@@

  \startcase{.2cm}
  \[
    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T_3}
    {
      \begin{array}{ccc}
        \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads \{t_1' \simeq t_2\}
        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
        & |t_1'| =_{\beta\eta} |t_1|
        \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [t_1/x] T' \tpsynth \star
        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpcheck [t_1/x]T'
        & [t_2/x]T' \cong T_3
      \end{array}
    }
  \]
  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t'
  \tpsynthleads \{t_1 \simeq t_2'\}\).
  By the IH on the second premise of the second row,
  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_1/x]\ T \tpsynth \star\)
  (\(\star\) is only convertible with itself).
  By the IH the second premise of the second row,
  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \tpcheck [t_1/x]T'\).
  By assumption, the third premise of the second row, and transitivity of
  convertibility, we have \([t_2/x]T' \cong T_4\) (\(T_4\) is the type we must
  check the entire expression against).
  We use the rule to conclude (\(\textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2) =
  \textit{dom}(\Gamma_1,X_1:\kappa_1',\Gamma_2)\)), noting that the measure is
  preserved.

@@

  \startcase{.2cm}
  \[
    \infer{
      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \cdot T' \tpsynth [T'/X]T
    }{
      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads
      \abs{\forall}{X}{\kappa}{T}
      & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T' \tpsynth \kappa'
      & \kappa'\cong\kappa
    }
  \]
  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t
  \tpsynthleads \abs{\forall}{X}{\kappa}{T}\).
  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T'
  \tpsynth \kappa''\) for some \(\kappa'' \cong \kappa'\), from which we obtain
  \(\kappa'' \cong \kappa\).
  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \cdot T' \tpsynth
  [T'/X]T\), and the measure is preserved.
\end{proof}

@@

  \startcase{.2cm}
  \[
    \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T_3}
    {
      \begin{array}{ccc}
        \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynthleads \{t_1' \simeq t_2\}
        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
        & |t_1'| =_{\beta\eta} |t_1|
        \\ \Gamma_1,x_1:T_1,\Gamma_2 \vdash [t_1/x] T' \tpsynth \star
        & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpcheck [t_1/x]T'
        & [t_2/x]T' \cong T_3
      \end{array}
    }
  \]
  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth
  T''\) (with no greater depth) for some \(T'' \cong \{t_1' \simeq t_2\}\).
  That means \(T'' \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
  such that \(|t_3| =_{\beta\eta} |t_1'|\) and \(|t_4| =_{\beta\eta} |t_2|\).
  So we further obtain that \(|t_3| =_{\beta\eta} |t_1|\).
  By the IH on the first premise of the second row, \(\Gamma_1,x_1:T_2,\Gamma_2
  \vdash [t_1/x]T \tpsynth \star\).
  By the IH on the second premise of the second row, \(\Gamma,x_1:T_2,\Gamma_2
  \vdash t \tpcheck [t_1/x]T'\).
  By assumption, the third premise of the second row, and transitivity of
  convertibility, we have \([t_2/x]T' \cong T_4\).
  We conclude with the rule, preserving the measure.
